---
title: Effects of Weather, Seasons, and Registration on Bike Rentals 
subtitle: "STAT 3302: Statistical Modeling for Discovery II"
author:
- Colin Farrell
- Paarth Mahajan
- Aayush Sharma
- Tony Zheng
group: "Group A (Bike Sharing)"
# Choose one format and delete the other:
format:
  revealjs:
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
  cache: true
  autodep: true
---

```{r}
#| label: setup
# Load packages and data here
library(tidyverse)
library(ggthemes) # for theme_clean

bike = read_csv("data/hour.csv")
```

## Introduction

<!-- Introduce the problem and your research question(s). Why is this
question interesting or important? -->

Bike rentals vary from day-to-day depending on various weather, temporal, and user conditions. By analyzing which conditions have the greatest effect on the estimated number of rentals, we can find what conditions have an impact or don't, and how these impacts interact with each other. As such, we ask the following questions:


## Research Questions

1. Does the fact that it is a workday (not a weekend/holiday) affect the number of Bike Rentals?

2. Do weather conditions affect the number of Bike Rentals among registered users vs. casual users?

3. Does the fact that it is a workday (or not) affect the number of Bike Rentals among registered users vs. casual users differently?

4. Do the number of rentals change through a season (e.g. do rentals taper off as a season progresses)?


# Methods

<!-- Describe the model you fit: response variable, distribution, link
function, and predictors. Justify your modeling choices (e.g., why you
included certain predictors, interactions, or polynomial terms). -->

## Model {auto-animate="true"}

We fit model
$$
Y_i \sim \text{Poisson}(\lambda_i)
$$

where $Y_i$ is the observed number of bike rentals for the hour of the $i$-th observation. For each hour, there are two $Y$:

- One for casual users
- One for registered users

And we have link function

$$
\log(\lambda_i) = X_i^T\beta
$$

## Model

Then, $X_i \in \mathbb{R}^{(p+1) \times n}$ and $\beta \in \mathbb{R}^{p + 1}$, for $i \in \{1, 2, \dots, n\}$

::: {.incremental}
- $n$ is the number of samples in the hour dataset times two (one for each registered count, one for casual)
- $p$ is the total number of predictors we're including, with intercept term $\beta_0 = \beta_{0, i}$ for $\forall i$
- $X_{i, 1} = 1$ for $\forall i$ 
- and all other $X_{i, j}$ correspond to the value $j$-th predictor for observation $i$
:::

## Predictors {.smaller}

- `temp`
- `windspeed`
- `hum`
- `weathersit` - Precipitation & cloud cover
- `season`
- `workingday`
- **User Type**
    - Indicator of if the number of rentals is taken from the `casual` or `registered` columns of the dataset (0 if `casual`, 1 if `registered`)
- **Interaction between `userType` and all other predictors**
- **Day of season** 
    - The day of the season (e.g. 0 for the first day of Spring, 1 for the second, etc..., then 0 for the first day of Fall, 1 for the second, etc...)

## Results

<!-- Present the fitted model, interpret key coefficients, and report
any hypothesis tests or confidence intervals. Use tables and figures
to support your findings. -->

```{r}
# data manipulation

# create UserType column
bike_transformed = bike %>% 
    pivot_longer( # split casual and registered users
        c("casual", "registered"),
        names_to = "user_type",
        values_to = "rentals"
    ) %>%
    group_by(yr, consecutive_id(season)) %>%
    mutate( # create day of season, create factors for weather/season
        day_of_season = as.numeric(dteday - min(dteday)),
        weathersit = factor(weathersit),
        season = factor(season, levels = c(1, 2, 3, 4), labels = c("Spr", "Sum", "Aut", "Win"))
    )
```

We get fitted model:
```{r}
model = glm(rentals ~ (temp + windspeed + hum + weathersit + workingday + day_of_season + season)*user_type, bike_transformed, family = poisson())

s = summary(model)
s$coefficients[, c("Estimate", "Pr(>|z|)")] |> round(digits = 4)
```

Which is the same after running `stepAIC` (backwards + forwards stepwise AIC model selection) starting from the full model and from the null model.

## Diagnostics

<!-- Describe at least one diagnostic check you performed (e.g.,
residual plots, goodness-of-fit tests). What did it reveal about
your model? -->

We also generate diagnostic plot for the estimated log rentals against residuals (for both Pearson and deviance residuals)

```{r}
residuals_data = bike_transformed |>
    ungroup() |>
    mutate(
        fitted.value = fitted.values(model),
        deviance.residuals = residuals(model, type = "deviance"),
        pearson.residuals = residuals(model, type = "pearson")
    )

# pearson residuals plot
ggplot(residuals_data, aes(x = fitted.value, y = pearson.residuals)) +
    geom_point(alpha = 0.1) +
    geom_smooth() +
    labs(
        title = "Pearson Residuals for Model",
        x = "Estimated # of Bike Rentals",
        y = "Pearson Residual"
    ) +
    theme_calc() +
    facet_wrap(~user_type)
```

## Diagnostics

We also generate diagnostic plot for the estimated log rentals against residuals (for both Pearson and deviance residuals)

```{r}
ggplot(residuals_data, aes(x = fitted.value, y = deviance.residuals)) +
    geom_point(alpha = 0.1) +
    geom_smooth() +
    labs(
        title = "Deviance Residuals for Model",
        x = "Estimated # of Bike Rentals",
        y = "Deviance Residual"
    ) +
    theme_calc() +
    facet_wrap(~user_type)
```

## (RQ.1)

Workdays also had an effect on the average number of rides, as the coefficient is statistically significant with associated p-value
```{r}
s$coefficients["workingday", "Pr(>|z|)"]
```

and the model is statistically significant compared to the model without both workday coefficients, having p-value on the two predictors
```{r}
model_no_workingday = glm(rentals ~ (temp + windspeed + hum + weathersit + day_of_season + season)*user_type, bike_transformed, family = poisson())

anova(model_no_workingday, model)[2, "Pr(>Chi)"]
```

## (RQ.1)

The coefficient value for the `workingday` predictor is
```{r}
coef(model)["workingday"]
```

Therefore, the estimated multiplicative change in mean number of riders when it is a working day compared to a non-working day, i.e., a holiday or weekend, is given by
```{r}
coef(model)[["workingday"]] |> exp()
```

which means that approximately 58% less average rides are estimated on a workday compared to a non-workday.

## (RQ.2)

Want to know: do weather conditions affect casual vs registered riders differently?

Weather Predictors Examined: temperature, humidity, wind speed, weather-situation. Also includes interactions between all predictors and `user_type`.

## (RQ.2)

```{r}
ggplot(bike_transformed, aes(x = temp, y = rentals, color = user_type)) +
    geom_point(alpha = 0.2) +
    labs(x = "Normalized Temperature", y = "Rentals", title = "Temperature vs Bike Rentals by User Type") +
    theme_clean()
```

## (RQ.2) {.smaller}

```{r}
library(kableExtra)
model_summary <- summary(model)$coefficients
weather_vars = c("temp", "hum", "windspeed", "weathersit2", "weathersit3", "weathersit4")
casual_coefs      <- model_summary[weather_vars, "Estimate"]
casual_pvals      <- model_summary[weather_vars, "Pr(>|z|)"]
interaction_terms <- paste0(weather_vars, ":user_typeregistered")
registered_coefs = model_summary[interaction_terms, "Estimate"]
interaction_pvals <- model_summary[interaction_terms, "Pr(>|z|)"]

weather_comparison <- data.frame(
    Variable      = c("Temperature", "Wind Speed", "Humidity",
                      "Weather: Cloudy/Mist", "Weather: Light Rain/Snow", "Weather: Heavy Rain/Snow"),
    Casual        = round(casual_coefs, 3),
    Registered    = round(registered_coefs, 3),
    Difference    = round(registered_coefs - casual_coefs, 3),
    P_Casual      = round(casual_pvals, 4),
    P_Interaction = round(interaction_pvals, 4)
)

weather_comparison |>
    kable(
        col.names = c("Variable", "Casual", "Registered", "Difference", "p (Casual)", "p (Interaction)"),
        caption   = "Weather Effects on Bike Rentals by User Type"
    ) |>
    kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## (RQ.2)

::::: r-stack
::: {.fragment .fade-in-then-out}
We can look at the corresponding coefficients for casual users vs registered users for each of the weather-related terms. For registered users, this would be the casual user coefficient plus the interaction term for registered users. All of `temp`, `hum`, and `windspeed` (and their respective interaction terms with `user_type`) were statistically significant, with low p-values, which means that these observed relationships were significant. This was also true for each of the seasons and their interactions terms, except for the baseline coefficient for heavy precipitation (though its interaction term was significant).
:::

::: {.fragment .fade-in-then-out}
Overall, causal users had strong sensitivity to temperature, humidity, and wind speed. Registered users are still affected by weather, but seems to be less affected, with the absolute value of each of their coefficients being smaller than the absolute value for casual users. Such a case is possibly due to more consistent baseline usage and behavior driven by necessity rather than comfort.
:::
:::::

## (RQ.3) {auto-animate="true"}

We want to know if workday status affects the bike rentals of casual and registered users differently.

In our model, the effect of `workingday` on different users (`user_type`) and the interaction effect is captured by these terms: 

To keep it concise let's use aliases for the parameter names for our fitted model. Let the main effects `wd = workingday`, `utr = user_typeregistered` and interaction effect `wd:utr = workin gday:user_typeregistered` :


$$
log(\lambda_i) = \cdots + \beta_{\text{wd}} \cdot \mathbf{1}[\text{wd}] + \beta_{\text{wd:utr}} \cdot \mathbf{1}[\text{wd}] \cdot \mathbf{1}[\text{utr}] + \beta_{utr} \cdot \mathbf{1}[\text{utr}]
$$

## (RQ.3) {auto-animate="true"}

$$
log(\lambda_i) = \cdots + \beta_{\text{wd}} \cdot \mathbf{1}[\text{wd}] + \beta_{\text{wd:utr}} \cdot \mathbf{1}[\text{wd}] \cdot \mathbf{1}[\text{utr}] + \beta_{utr} \cdot \mathbf{1}[\text{utr}]
$$

To answer our question we look at the interaction term $\beta_{\text{workingday:registered}}$  which captures this difference. The p-value generated for this interaction effect during model fit:

\

```{r}
#Get cofficient names
coefs <- c("workingday", "workingday:user_typeregistered", "user_typeregistered")

#p-values
s$coefficients[coefs, c("Estimate", "Std. Error", "Pr(>|z|)")]
```


## (RQ.3) {auto-animate="true"}


The interaction term `wd:utr` is statistically significant since (p < 0.05), suggesting that we enough evidence to reject the idea that the effect of whether it is a working day on Bike Rentals differs by user type considering everything other predictor is held at the same value.


```{r}
#Get cofficient names
coefs <- c("workingday", "workingday:user_typeregistered", "user_typeregistered")

#p-values
s$coefficients[coefs, c("Estimate", "Std. Error", "Pr(>|z|)")]
```

## (RQ.3) {auto-animate="true"}

We can confirm this by calculating the difference in Rentals for non-working/working days for different user type:

```{r}
#Coeffs
wday_coefs <- c("workingday", "workingday:user_typeregistered")

# Multiplicative effect on mean rentals
casual_mult    <- exp(coef(model)[["workingday"]])
registered_mult <- exp(sum(coef(model)[wday_coefs]))

cat("Casual Multiplicative Difference in Rentals for Workingday:   ", round(casual_mult, 3), "\n")
cat("Registered Multiplicative Difference in Rentals for Workingday:", round(registered_mult, 3), "\n")
```

$$
\frac{\lambda_{wd,cas}}{\lambda_{nw,cas}} = exp\{ \beta_{wd}\} = 0.42
$$


$$
\frac{\lambda_{wd,reg}}{\lambda_{nw,reg}} = exp\{ \beta_{wd} + \beta_{wd:utr} \} = 1.313
$$

## (RQ.3) {auto-animate="true"}

$$
\frac{\lambda_{wd,cas}}{\lambda_{nw,cas}} = exp\{ \beta_{wd}\} = 0.42
$$


$$
\frac{\lambda_{wd,reg}}{\lambda_{nw,reg}} = exp\{ \beta_{wd} + \beta_{wd:utr} \} = 1.313
$$

On workdays, casual rentals are estimated to change by a factor of 0.42 while registered rentals change by a factor of 1.313 — these indicate a reversal in the direction of the change in the rate rentals.

## (RQ.3) {auto-animate="true"}

We can visualize this difference by plotting the Estimated Hourly Rentals for different types of days for some median weather conditions (to hold those parameters constant)

```{r}
# Predicted mean rentals at median weather conditions by user_type × working day
pred_data <- bike_transformed |>
  ungroup() |>
  group_by(user_type, workingday) |>
  summarise(
    temp       = median(temp),
    windspeed  = median(windspeed),
    hum        = median(hum),
    weathersit = factor(1, levels = levels(weathersit)),
    season     = factor("Sum", levels = levels(season)),
    day_of_season = median(day_of_season),
    .groups = "drop"
  ) |>
  mutate(predicted = predict(model, newdata = pick(everything()), type = "response"))

ggplot(pred_data, aes(x = user_type, y = predicted, fill = factor(workingday,
       labels = c("Non-Workday", "Workday")))) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("#619CFF", "#F8766D"), name = "Day Type") +
  labs(
    title    = "Estimated Mean Hourly Rentals: Workday vs Non-Workday",
    subtitle = "At median weather conditions (Summer, weathersit = 1)",
    x        = "User Type",
    y        = "Predicted Mean Rentals"
  ) +
  theme_calc()
```

## (RQ.4)

```{r}
ggplot(bike_transformed, aes(x = day_of_season, y = rentals, color = user_type)) +
    geom_point(alpha = 0.1) +
    labs(title = "# of Rentals vs. Day of Season") +
    theme_clean()
```

## (RQ.4)

Our estimated `day_of_season` coefficients were statistically significant, with individual p-values
```{r}
usertype_coefs = c("day_of_season", "day_of_season:user_typeregistered")
s$coefficients[usertype_coefs, "Pr(>|z|)"]
```

and adding both `day_of_season` and `day_of_season:user_typeregistered` to the model had p-value of 
```{r}
no_day_of_season_model = glm(rentals ~ (temp + windspeed + hum + weathersit + workingday + season)*user_type, bike_transformed, family = poisson())

anova(no_day_of_season_model, model)[2, "Pr(>Chi)"]
```

with individual values
```{r}
coef(model)[usertype_coefs]
```

## (RQ.4)

```{r}
casual_dos_coef = coef(model)[["day_of_season"]]
reg_dos_coef = coef(model)[usertype_coefs] |> sum()
```

There was a significant positive relationship: for every day that passed in a season, the estimated mean number of hourly casual bike rentals increased for by approx. factor of...


$\exp( {`r casual_dos_coef |> format(digits = 5)`} ) \approx `r casual_dos_coef |> exp() |> signif(digits = 5)`$ for **casual users** 

$\exp( {`r reg_dos_coef |> format(digits = 5)`} ) \approx `r reg_dos_coef |> exp() |> format(digits = 5)`$ for **registered users**

## Conclusion

<!-- Briefly summarize your main findings and their implications. -->
Bike rentals are significantly associated with different climate/weather conditions. They...

::: {.incremental}
- Significantly increased for higher temperatures
- Increased for non-working days
- Increased as a season went along (later parts of a season were associated with more rentals)
:::
