---
title: Effect of Weather, Day of Season, and Registration Status on Bike Rentals 
subtitle: "STAT 3302: Statistical Modeling for Discovery II"
author:
- Colin Farrell
- Paarth Mahajan
- Aayush Sharma
- Tony Zheng
group: "Group A (Bike Sharing)"
date: today
# Choose one format and delete the other:
format:
  html:
    embed-resources: true
  pdf:
    geometry:
      - margin=1in
    fontsize: 11pt
bibliography: dataset.bib
execute:
  echo: false
  warning: false
  message: false
  cache: true
  autodep: true
---

```{r}
#| label: setup
# Load packages and data here
library(tidyverse)
library(ggthemes) # for theme_clean

bike = read_csv("data/hour.csv")
```

## Abstract

<!-- 2-3 sentence summary of your question, method, and key finding. -->

## Introduction

<!-- Introduce the problem and your research question(s). Why is this
question interesting or important? -->

Bike rentals vary from day-to-day depending on various weather, temporal, and user conditions. By analyzing which conditions have the greatest effect on the estimated number of rentals, we can find what conditions have an impact or don't, and how these impacts interact with each other. As such, we ask the following questions:

#### (RQ.1)
Does the temperature of a day affect the number of Bike Rentals after accounting for other weather conditions (e.g. precipitation, humidity, and wind speed), user type and type of day (workday/not)?

#### (RQ.2)
Does the fact that it is a workday (not a weekend/holiday) affect the number of Bike Rentals after accounting for weather conditions (e.g. precipitation, humidity, temperature, and wind speed), and user type?

#### (RQ.3)
Do weather conditions affect the number of Bike Rentals differently between registered users vs. casual users?

#### (RQ.4)
Does the fact that it is a workday (or not) affect the number of Bike Rentals among registered users vs. casual users differently?

#### (RQ.5)
Do the number of rentals change through a season (e.g. do rentals taper of as a season progresses)?

These questions address individual bike rental conditions and the way user they interact between each other (if one condition occurring cancels out another or strengthens its effect).

## Data

<!-- Describe the dataset: source, unit of observation, sample size,
response variable, and key predictors. Include relevant summary
statistics or exploratory plots. -->

The bike sharing dataset was [@bike_sharing_275]

```{r}
# data manipulation

# create UserType column
bike_transformed = bike %>% 
    pivot_longer( # split casual and registered users
        c("casual", "registered"),
        names_to = "user_type",
        values_to = "rentals"
    ) %>%
    group_by(yr, consecutive_id(season)) %>%
    mutate( # create day of season, create factors for weather/season
        day_of_season = as.numeric(dteday - min(dteday)),
        weathersit = factor(weathersit),
        season = factor(season, levels = c(1, 2, 3, 4), labels = c("Spr", "Sum", "Aut", "Win"))
    )
```

### Exploratory Plots
```{r}
ggplot(bike_transformed, aes(x = temp, y = rentals, color = user_type)) +
    geom_point(alpha = 0.2) +
    labs(x = "Normalized Temperature", y = "Rentals", title = "Temperature vs Bike Rentals by User Type") +
    theme_clean()
```

## Methods

<!-- Describe the model you fit: response variable, distribution, link
function, and predictors. Justify your modeling choices (e.g., why you
included certain predictors, interactions, or polynomial terms). -->

### Model

We fit model
$$
Y_i \sim \text{Poisson}(\lambda_i)
$$

with link function
$$
\log(\lambda_i) = X_i^T\beta
$$

for $i \in \{1, 2, \dots, n\}$.

$Y_i$ is the observed number of bike rentals for the hour of the $i$-th observation. For each hour, there are two $Y$:

- One for casual users
- One for registered users

Then, $X_i \in \mathbb{R}^{(p+1) \times n}$ and $\beta \in \mathbb{R}^{p + 1}$, where
- $n$ is the number of samples in the hour dataset times two (one for each registered count, one for casual)
- $p$ is the total number of predictors we're including, with intercept term $\beta_0 = \beta_{0, i}$ for $\forall i$
- $X_{i, 1} = 1$ for $\forall i$ 
- and all other $X_{i, j}$ correspond to the value $j$-th predictor for observation $i$ $i \in {1, 2, \dots, n}$.

### Predictors

Our model will include:

| Predictor | Description | Reason |
| ----- | ------ | ----------- |
| `temp` | The temperature of a day in ($^\circ F$) | To control for the effect of the different aspects of the weather on the number of bike rentals |
| `windspeed` | The normalized wind speed from 0-67 (unitless) |
| `hum` | The normalized humidity from 0-100 (unitless) |
| `weathersit` | Factor representing the amount of precipitation/cloud cover. In $\{1, 2, 3, 4\}$, where <br> **1:** Clear, Few clouds, Partly cloudy, Partly cloudy <br> **2:** Mist + Cloudy, Mist + Broken clouds, Mist + Few clouds, Mist <br> **3:** Light Snow, Light Rain + Thunderstorm + Scattered clouds, Light Rain + Scattered clouds <br> **4:** Heavy Rain + Ice Pallets + Thunderstorm + Mist, Snow + Fog | To examine the effect of the amount of precipitation on the number of bike rentals, after accounting for other parts of the weather |
| User Type | Indicator of if the number of rentals is taken from the `casual` or `registered` columns of the dataset (0 if `casual`, 1 if `registered`) | Included to account for the effect of being a registered user by itself on rentals |
| `season` | Indicator for what season it is. In the data, is an integer from 1-4, where 1 is Spring, 2 Summer, 3 Fall, and 4 Winter. For our model, we transform it into categorical data where `Spr` is for Spring, `Sum` is for Summer, `Aut` is for Fall, and `Win` is for Winter | Included to account for the effect of the season on bike rentals |
| `workingday`| Indicator of if day was a workday or not (1 if workday, 0 if not) | Helps identify a workday and accounts for the effect of that fact on the number of rentals for the given hour |
| Day of season | The day of the season (e.g. 0 for the first day of Spring, 1 for the second, etc..., then 0 for the first day of Fall, 1 for the second, etc...). | Helps to answer [RQ.5](#rq.5) by showing if there is a significant linear relation between how long it's been a season and the number of rentals. |
| Interaction between `userType` and all other predictors | | To see how registered users are affected by each other predictor compared to casual users |

We also use `stepAIC` (bidirectional stepwise model selection using AIC) to determine if our model would have a better fit by ablating any predictors.

## Results

<!-- Present the fitted model, interpret key coefficients, and report
any hypothesis tests or confidence intervals. Use tables and figures
to support your findings. -->

We get fitted model:
```{r}
#| label: model_fit

model = glm(rentals ~ (temp + windspeed + hum + weathersit + workingday + day_of_season + season)*user_type, bike_transformed, family = poisson())

s = summary(model)
s
```

Which is the same after running `stepAIC` (backwards + forwards stepwise AIC model selection) starting from the full model and from the null model.
```{r, eval=FALSE}
step_model_full = MASS::stepAIC(model, trace = 0, direction = "both")
step_model_full
```

```{r, eval=FALSE}
null_model = glm(rentals ~ 1, bike_transformed, family = poisson())
step_model_null = MASS::stepAIC(null_model, trace = 0, scope = list(lower = null_model, upper = model, direction = "both"))
step_model_null
```

## Diagnostics

<!-- Describe at least one diagnostic check you performed (e.g.,
residual plots, goodness-of-fit tests). What did it reveal about
your model? -->

We can generate a diagnostic plot for the estimated log rentals against residuals (for both Pearson and deviance residuals)

```{r}
residuals_data = bike_transformed |>
    ungroup() |>
    mutate(
        fitted.value = fitted.values(model),
        deviance.residuals = residuals(model, type = "deviance"),
        pearson.residuals = residuals(model, type = "pearson")
    )

# pearson residuals plot
ggplot(residuals_data, aes(x = fitted.value, y = pearson.residuals)) +
    geom_point(alpha = 0.1) +
    geom_smooth() +
    labs(
        title = "Pearson Residuals for Model",
        x = "Estimated # of Bike Rentals",
        y = "Pearson Residual"
    ) +
    theme_calc() +
    facet_wrap(~user_type)

# deviance residuals plot
ggplot(residuals_data, aes(x = fitted.value, y = deviance.residuals)) +
    geom_point(alpha = 0.1) +
    geom_smooth() +
    labs(
        title = "Deviance Residuals for Model",
        x = "Estimated # of Bike Rentals",
        y = "Deviance Residual"
    ) +
    theme_calc() +
    facet_wrap(~user_type)
```

It appears that our model's residuals are fairly well-fitted for low to medium amounts of bike rentals (0 to about 200). They seem to be centered around 0 for both Pearson and deviance.

For larger values, it appears that both types of residual are biased negatively, which means that our model is sometimes overestimating the number of rentals. This means that our model might not be appropriate; possible causes might be that there is some factor that causes bike rentals to be lower in these cases, or that our predictors are not truly linear (e.g. it might be that a more accurate relationship would be that $\log(\mathbb{E}(rentals))$ is proportional to the log of temperature, rather than the actual temperature).

## Discussion

<!-- Discuss your findings in context. What are the limitations of
your analysis? What would you do differently with more time or data? -->

### (RQ.1)

The estimated coefficients for `temp` are seen to be statistically significant, as the individual p-values for `temp` and `temp:user_typeregistered` are given by:
```{r}
temp_usertype_coefs <- c("temp", "temp:user_typeregistered")
s$coefficients[temp_usertype_coefs, "Pr(>|z|)"]
```
with combined p-value
```{r}
model_no_temp = glm(rentals ~ (windspeed + hum + weathersit + workingday + day_of_season + season)*user_type, bike_transformed, family = poisson())
anova(model_no_temp, model)[2, "Pr(>Chi)"]
```

The individual values of the coefficients are 
```{r}
coef(model)[temp_usertype_coefs]
```

meaning for casual users, a one degree increase in temperature results in an estimated multiplicative increase in mean bike use by
```{r}
coef(model)[["temp"]] |> exp()
```

and, for casual users, a smaller yet significant estimated mean increase of
```{r}
coef(model)[temp_usertype_coefs] |> sum() |> exp()
```


### (RQ.2)

Workdays also had an effect on the average number of rides, as the coefficient is statistically significant with associated p-value
```{r}
s$coefficients["workingday", "Pr(>|z|)"]
```

and the model is statistically significant compared to the model without both workday coefficients, having p-value on the two predictors
```{r}
model_no_workingday = glm(rentals ~ (temp + windspeed + hum + weathersit + day_of_season + season)*user_type, bike_transformed, family = poisson())

anova(model_no_workingday, model)[2, "Pr(>Chi)"]
```

The coefficient value for the `workingday` predictor is
```{r}
coef(model)["workingday"]
```

Therefore, the estimated multiplicative change in mean number of riders when it is a working day compared to a non-working day, i.e., a holiday or weekend, is given by
```{r}
coef(model)[["workingday"]] |> exp()
```

which means that approximately 58% less average rides are estimated on a workday compared to a non-workday.

### (RQ.3)

Does Weather affect casual users differently than registered users?

Coefficient summary for all weather conditions:
```{r}
weather_coefs = c(grepl("weathersit|temp|windspeed|hum", names(coef(model))))
s$coefficients[weather_coefs, c("Estimate", "Pr(>|z|)")]
```

Weather conditions affect both groups of users, but casual users are substantially more sensitive to weather than registered users. For example, the estimated mean rentals registered users are less affected by the temperature: the coefficient for temperature (i.e. the factor by which the estimated log mean number of rentals increases for unit of normalized temperature) for casual users is `r coef(model)[["temp"]] |> round(digits = 2)` vs. (`r  coef(model)[["temp"]] |> round(digits = 2)` + `r coef(model)[["temp:user_typeregistered"]] |> round(digits = 2)` = `r coef(model)[c("temp", "temp:user_typeregistered")] |> round(digits = 2) |> sum()`) for registered users.

The same goes for humidity, with estimated casual humidity coefficient `r coef(model)[["hum"]] |> round(digits = 2)` vs. registered humidity coefficient (`r coef(model)[["hum"]] |> round(digits = 2)` + `r coef(model)[["hum:user_typeregistered"]] |> round(digits = 2)` = `r coef(model)[c("hum", "hum:user_typeregistered")] |> round(digits = 2) |> sum()`).

For windspeed, the estimated casual coefficient was `r coef(model)[["windspeed"]] |> round(digits = 2)` vs. registered windspeedidity coefficient (`r coef(model)[["windspeed"]] |> round(digits = 2)` + `r coef(model)[["windspeed:user_typeregistered"]] |> round(digits = 2)` = `r coef(model)[c("windspeed", "windspeed:user_typeregistered")] |> round(digits = 2) |> sum()`).

Finally, for precipitation, the casual coefficients were
```{r}
coef(model)[grepl("weathersit\\d$", names(coef(model)))]
```

versus
```{r}
coef(model)[grepl("weathersit\\d$)", names(coef(model)))] + coef(model)[grepl("weathersit\\d:user_typeregistered", names(coef(model)))]
```

All of `temp`, `hum`, and `windspeed` (and their respective interaction terms with `user_type`) were statistically significant, with low p-values, which means that these observed relationships were significant. This was also true for each of the seasons and their interactions terms, except for the baseline coefficient for heavy precipitation (though its interaction term was significant). This indicates that there was not a statistically significant difference in bike rentals among casual users between a heavy precipitation versus. clear day, though this could be due to high collinearity between such conditions and other weather conditions (like windspeed and humidity).

Overall, causal users had strong sensitivity to temperature, humidity, and windspeed. Registered users are still affected by weather, but seems to be less affected, with the absolute value of each of their coefficients being smaller than the absolute value for casual users. Such a case is possibly due to more consistent baseline usage and behavior driven by necessity rather than comfort. 

### (RQ.5)

Our estimated `day_of_season` coefficients were statistically significant, since `day_of_season` and `day_of_season:user_typeregistered` had individual p-values of
```{r}
usertype_coefs = c("day_of_season", "day_of_season:user_typeregistered")
s$coefficients[usertype_coefs, "Pr(>|z|)"]
```

and a p-value when added together
```{r}
no_day_of_season_model = glm(rentals ~ (temp + windspeed + hum + weathersit + workingday + season)*user_type, bike_transformed, family = poisson())

anova(no_day_of_season_model, model)[2, "Pr(>Chi)"]
```

with individual values
```{r}
coef(model)[usertype_coefs]
```

These indicate that there was a significant positive relationship between the day of the season and average bike rentals. Bike rentals increased in later parts of seasons for all users - though to a slightly greater extent for registered users. Specifically, it is estimated that for every day that passed in a season, the estimated mean number of casual bike rentals in an hour increased by a factor of 
```{r}
coef(model)[["day_of_season"]] |> exp()
```
and for registered users
```{r}
coef(model)[usertype_coefs] |> sum() |> exp()
```

## Conclusion

<!-- Briefly summarize your main findings and their implications. -->

Bike rentals are significantly associated with different climate/weather conditions. They

- Significantly increased for higher temperatures
- Increased for non-working days
- Increased as a season went along (later parts of a season were associated with more rentals)

## References

<!-- Cite the data source and any articles or resources you used. -->

::: {#refs}
:::

## Appendix {.appendix}

<!-- Optional: include additional code, tables, or figures here.
This section does not count toward the page limit. -->

The model was fit using

```{r, echo=TRUE, eval=FALSE, ref.label="model_fit"}
#| code-overflow: wrap
```

