---
title: Effect of Weather, Day of Season, and Registration Status on Bike Rentals 
subtitle: "STAT 3302: Statistical Modeling for Discovery II"
author:
- Colin Farrell
- Paarth Mahajan
- Aayush Sharma
- Tony Zheng
group: "Group A (Bike Sharing)"
date: today
# Choose one format and delete the other:
format:
  html:
    embed-resources: true
  pdf:
    geometry:
      - margin=1in
    fontsize: 11pt
bibliography: dataset.bib
execute:
  echo: false
  warning: false
  message: false
  cache: true
  autodep: true
---

```{r}
#| label: setup
# Load packages and data here
library(tidyverse)
library(ggthemes) # for theme_clean

bike = read_csv("data/hour.csv")
```

## Abstract

<!-- 2-3 sentence summary of your question, method, and key finding. -->

There are many factors that depend why someone would use a bike-sharing service. We believe that it is of interest to see which factors impact bike rental numbers the most, specifically looking at whether or not it is a workday, the weather conditions, and the season, also breaking down the response into registered users who pay a subscription to the service and casual users who are picking one up on a one-time basis. We fit a Poisson regression model with a log-rate link function, with the following findings: users typically ride on non-working days overall, but registered users are more likely to ride on workdays; registered users are typically affected less by weather; and the progression of a season does not typically result with an increase of total riders.

## Introduction

<!-- Introduce the problem and your research question(s). Why is this
question interesting or important? -->

Bike rentals vary from day-to-day depending on various weather, temporal, and user conditions. By analyzing which conditions have the greatest effect on the estimated number of rentals, we can find what conditions have an impact or don't, and how these impacts interact with each other.

#### (RQ.1)
Does the fact that it is a workday (not a weekend/holiday) affect the number of Bike Rentals?

#### (RQ.2)
Do weather conditions affect the number of Bike Rentals among registered users vs. casual users?

#### (RQ.3)
Does the fact that it is a workday (or not) affect the number of Bike Rentals among registered users vs. casual users differently?

#### (RQ.4)
Do the number of rentals change through a season (e.g. do rentals taper off as a season progresses)?


## Data

<!-- Describe the dataset: source, unit of observation, sample size,
response variable, and key predictors. Include relevant summary
statistics or exploratory plots. -->

The Bike Sharing dataset was sourced from [@bike_sharing_275]. It contains the daily and hourly counts of bike rentals in the Capital bike-share system in 2011 and 2012, augmented with weather and temporal information. We use the hourly dataset; it contains a total of `r nrow(bike)` counts.

### Predictors

Dataset predictors used in model fitting: (check appendix for expanded details)

| Predictor | Description |
| -- | ------ |
| `temp` | The normalized temperature of the hour (unitless), in $\left[0, 1\right]$ |
| `windspeed` | The normalized windspeed of a day (unitless), in $\left[0, 1\right]$ |
| `hum` | The normalized humidity (unitless), in $\left[0, 1\right]$|
| `weathersit` | Factor representing the amount/degree of precipitation/cloud cover |
| `season` | Indicator for Season |
| `workingday`| Indicator of if day was a workday or not (1 if workday, 0 if not) |

### Created Predictors

New predictors from the dataset, for model fit:

| Predictor | Description |
| -- | ------ |
| `user_type` | Indicator of if the number of rentals is taken from the `casual` or `registered` columns of the dataset (0 if `casual`, 1 if `registered`) |
| Day of season | The day of the season (e.g. 0 for the first day of Spring) |

```{r}
# data manipulation

# create UserType column
bike_transformed = bike %>% 
    pivot_longer( # split casual and registered users
        c("casual", "registered"),
        names_to = "user_type",
        values_to = "rentals"
    ) %>%
    group_by(yr, consecutive_id(season)) %>%
    mutate( # create day of season, create factors for weather/season
        day_of_season = as.numeric(dteday - min(dteday)),
        weathersit = factor(weathersit),
        season = factor(season, levels = c(1, 2, 3, 4), labels = c("Spr", "Sum", "Aut", "Win"))
    )
```

Our modified dataset then contains `r nrow(bike_transformed)` counts (i.e. double that of the original, since each row gets split into two).

Our model also includes the interaction term between all those described in [Predictors](#predictors) and `userType`. This is done to see how registered users are affected by each other predictor compared to casual users.

### Exploratory Plots

#### By `workday`

We can look at how [RQ.1](#rq.1) and [RQ.3](#rq.3) by plotting `workingday` against the rental counts, faceted by both `user_type` and `workingday`. We plot the proportion of counts accounted for by each bin of rentals in each group to see how the distribution itself changes, without it being affected by the total amount of counts in each group.

```{r}
ggplot(bike_transformed, aes(x = rentals)) +
    geom_histogram(aes(y = after_stat(density * width) / sum(after_stat(density * width)))) +
    facet_grid(user_type ~ factor(workingday, levels = c(0, 1), labels = c("Non-working Day", "Working Day"))) +
    labs(y = "Proportion of Counts, for user_type & workingday group") +
    theme_clean()
```

It seems that we generally see a significant change in rentals on working days, but this association is **reversed** for registered users. Casual users tend to see higher rental counts on *non-working* days, but registered users tend to see higher rental counts on *working* days. This suggests that the answer to [RQ.3](#rq.3) might be that registered and casual users **are** affected by working days differently.

#### By `user_type`

```{r}
ggplot(bike_transformed, aes(x = rentals)) +
    geom_histogram() +
    labs(title = "Bike Rentals by User Type") +
    facet_grid(user_type ~ .) +
    theme_clean()
```

It seems that the distribution of registered rentals is skewed higher than for casual users: casual counts are higher when the # of rentals is less than about 100, but registered counts are higher for rentals greater than that. In other words, it seems that `user_type` has an impact on the total number of rentals.

#### By `temp`

```{r}
ggplot(bike_transformed, aes(x = temp, y = rentals, color = user_type)) +
    geom_point(alpha = 0.2) +
    labs(x = "Normalized Temperature", y = "Rentals", title = "Temperature vs Bike Rentals by User Type") +
    theme_clean()
```

We can see that registered users generally have a higher baseline usage from casual users, even when comparing rentals for registered users at lower temperatures to casual users at higher temperatures.

We can also see that there seems to be a loosely positive association between temperature and the number of rentals; as the temperature increased, the number of rentals seemed to increase. However, it doesn't seem that there is a large difference between registered and casual users in the size of this association (i.e. the amount of rentals increases about the same per temperature increase between casual and registered users).

#### By `day_of_season`

We can explore [RQ.4](#rq.4) by plotting the `day_of_season` against the rental counts.

```{r}
ggplot(bike_transformed, aes(x = day_of_season, y = rentals, color = user_type)) +
    geom_point(alpha = 0.1) +
    labs(title = "# of Rentals vs. Day of Season") +
    theme_clean()
```

It doesn't seem that there's a strong relationship here; regardless of how far a season is along, it seems that the distribution of rentals stays relatively constant.

## Methods

<!-- Describe the model you fit: response variable, distribution, link
function, and predictors. Justify your modeling choices (e.g., why you
included certain predictors, interactions, or polynomial terms). -->

### Model

We fit model
$$
Y_i \sim \text{Poisson}(\lambda_i)
$$

with the log-rate link function
$$
\log(\lambda_i) = X_i^T\beta
$$

for $i \in \{1, 2, \dots, n\}$.

$Y_i$ is the observed number of bike rentals for the hour of the $i$-th observation. For each hour, there are two $Y$ based on the type of user (registered/casual).

Then, $X_i \in \mathbb{R}^{(p+1)}$ and $\beta \in \mathbb{R}^{p + 1}$, where:

- $n$ is the number of samples in the **hour dataset** times two (for each type of user)

- $p$ is the total number of predictors we're including

- $X_{i, j}$ correspond to the value $j$-th predictor for observation $i$, where $X_{i,1}=1$ for all $i$

We also use `stepAIC` (bidirectional stepwise model selection using AIC) to determine if our model would have a better fit by ablating any predictors.

## Results

<!-- Present the fitted model, interpret key coefficients, and report
any hypothesis tests or confidence intervals. Use tables and figures
to support your findings. -->

We get fitted model:
```{r}
#| label: model_fit

model = glm(rentals ~ (temp + windspeed + hum + weathersit + workingday + day_of_season + season)*user_type, bike_transformed, family = poisson())

s = summary(model)
s
```

Which is the same after running `stepAIC` (backwards + forwards stepwise AIC model selection) starting from the full model and from the null model.
```{r, eval=FALSE}
step_model_full = MASS::stepAIC(model, trace = 0, direction = "both")
step_model_full
```

```{r, eval=FALSE}
null_model = glm(rentals ~ 1, bike_transformed, family = poisson())
step_model_null = MASS::stepAIC(null_model, trace = 0, scope = list(lower = null_model, upper = model, direction = "both"))
step_model_null
```

## Diagnostics

<!-- Describe at least one diagnostic check you performed (e.g.,
residual plots, goodness-of-fit tests). What did it reveal about
your model? -->

We can generate a diagnostic plot for the estimated log rentals against residuals (for both Pearson and deviance residuals)

```{r}
residuals_data = bike_transformed |>
    ungroup() |>
    mutate(
        fitted.value = fitted.values(model),
        deviance.residuals = residuals(model, type = "deviance"),
        pearson.residuals = residuals(model, type = "pearson")
    )

# pearson residuals plot
ggplot(residuals_data, aes(x = fitted.value, y = pearson.residuals)) +
    geom_point(alpha = 0.1) +
    geom_smooth() +
    labs(
        title = "Pearson Residuals for Model",
        x = "Estimated # of Bike Rentals",
        y = "Pearson Residual"
    ) +
    theme_calc() +
    facet_wrap(~user_type)

# deviance residuals plot
ggplot(residuals_data, aes(x = fitted.value, y = deviance.residuals)) +
    geom_point(alpha = 0.1) +
    geom_smooth() +
    labs(
        title = "Deviance Residuals for Model",
        x = "Estimated # of Bike Rentals",
        y = "Deviance Residual"
    ) +
    theme_calc() +
    facet_wrap(~user_type)
```

It appears that our model's residuals are fairly well-fitted for low to medium amounts of bike rentals (0 to about 200). They seem to be centered around 0 for both Pearson and deviance.

For larger values, it appears that both types of residual are biased negatively, which means that our model is sometimes overestimating the number of rentals. This means that our model might not be appropriate; possible causes might be that there is some factor that causes bike rentals to be lower in these cases, or that our predictors are not truly linear (e.g. it might be that a more accurate relationship would be that $\log(\mathbb{E}(rentals))$ is proportional to the log of temperature, rather than the actual temperature).

## Discussion

<!-- Discuss your findings in context. What are the limitations of
your analysis? What would you do differently with more time or data? -->

### (RQ.1)

Workdays had an effect on the average number of rides, as the coefficient is statistically significant with associated p-value
```{r}
s$coefficients["workingday", "Pr(>|z|)"]
```

and the model is statistically significant compared to the model without both workday coefficients, having p-value on the two predictors
```{r}
model_no_workingday = glm(rentals ~ (temp + windspeed + hum + weathersit + day_of_season + season)*user_type, bike_transformed, family = poisson())

anova(model_no_workingday, model)[2, "Pr(>Chi)"]
```

The coefficient value for the `workingday` predictor is
```{r}
coef(model)["workingday"]
```

Therefore, the estimated multiplicative change in mean number of riders when it is a working day compared to a non-working day, i.e., a holiday or weekend, is given by
```{r}
coef(model)[["workingday"]] |> exp()
```

which means that approximately 58% less average rides are estimated on a workday compared to a non-workday.

### (RQ.2)

Does Weather affect casual users differently than registered users?

Coefficient summary for all weather conditions:
```{r}
weather_coefs = c(grepl("weathersit|temp|windspeed|hum", names(coef(model))))
s$coefficients[weather_coefs, c("Estimate", "Pr(>|z|)")]
```

Weather conditions affect both groups of users, but casual users are substantially more sensitive to weather than registered users. For example, the estimated mean rentals registered users are less affected by the temperature: the coefficient for temperature (i.e. the factor by which the estimated log mean number of rentals increases for unit of normalized temperature) for casual users is `r coef(model)[["temp"]] |> round(digits = 2)` vs. (`r  coef(model)[["temp"]] |> round(digits = 2)` + `r coef(model)[["temp:user_typeregistered"]] |> round(digits = 2)` = `r coef(model)[c("temp", "temp:user_typeregistered")] |> round(digits = 2) |> sum()`) for registered users.

The same goes for humidity, with estimated casual humidity coefficient `r coef(model)[["hum"]] |> round(digits = 2)` vs. registered humidity coefficient (`r coef(model)[["hum"]] |> round(digits = 2)` + `r coef(model)[["hum:user_typeregistered"]] |> round(digits = 2)` = `r coef(model)[c("hum", "hum:user_typeregistered")] |> round(digits = 2) |> sum()`).

For windspeed, the estimated casual coefficient was `r coef(model)[["windspeed"]] |> round(digits = 2)` vs. registered windspeedidity coefficient (`r coef(model)[["windspeed"]] |> round(digits = 2)` + `r coef(model)[["windspeed:user_typeregistered"]] |> round(digits = 2)` = `r coef(model)[c("windspeed", "windspeed:user_typeregistered")] |> round(digits = 2) |> sum()`).

Finally, for precipitation, the casual coefficients were
```{r}
coef(model)[grepl("weathersit\\d$", names(coef(model)))]
```

versus
```{r}
coef(model)[grepl("weathersit\\d$)", names(coef(model)))] + coef(model)[grepl("weathersit\\d:user_typeregistered", names(coef(model)))]
```

All of `temp`, `hum`, and `windspeed` (and their respective interaction terms with `user_type`) were statistically significant, with low p-values, which means that these observed relationships were significant. This was also true for each of the seasons and their interactions terms, except for the baseline coefficient for heavy precipitation (though its interaction term was significant). This indicates that there was not a statistically significant difference in bike rentals among casual users between a heavy precipitation versus. clear day, though this could be due to high collinearity between such conditions and other weather conditions (like windspeed and humidity).

Overall, causal users had strong sensitivity to temperature, humidity, and windspeed. Registered users are still affected by weather, but seems to be less affected, with the absolute value of each of their coefficients being smaller than the absolute value for casual users. Such a case is possibly due to more consistent baseline usage and behavior driven by necessity rather than comfort. 

### (RQ.3)


We want to know if workday status affects the bike rentals of casual and registered users differently.

The estimates for this interaction effect coefficient during model fit:

```{r}
#Get cofficient names
coefs <- c("workingday", "workingday:user_typeregistered", "user_typeregistered")

#p-values
s$coefficients[coefs, c("Estimate", "Std. Error", "Pr(>|z|)")]

#Coeffs
wday_coefs <- c("workingday", "workingday:user_typeregistered")

# Multiplicative effect on mean rentals
casual_mult    <- exp(coef(model)[["workingday"]])
registered_mult <- exp(sum(coef(model)[wday_coefs]))

cat("Casual Multiplicative Difference in Rentals for Workingday:   ", round(casual_mult, 3), "\n")
cat("Registered Multiplicative Difference in Rentals for Workingday:", round(registered_mult, 3), "\n")
```

We illustrate the multiplicative difference in rentals between Working and non-Working days for Casual and Registered users above. We visualize this below: 

```{r}
library(patchwork)

# --- Shared prediction data (your existing code, unchanged) ---
pred_data <- bike_transformed |>
  ungroup() |>
  group_by(user_type, workingday) |>
  summarise(
    temp          = median(temp),
    windspeed     = median(windspeed),
    hum           = median(hum),
    weathersit    = factor(1, levels = levels(weathersit)),
    season        = factor("Sum", levels = levels(season)),
    day_of_season = median(day_of_season),
    .groups = "drop"
  ) |>
  mutate(
    predicted = predict(model, newdata = pick(everything()), type = "response"),
    day_label = factor(workingday, labels = c("Non-Workday", "Workday"))
  )

# --- Plot 1: Your existing bar chart ---
p1 <- ggplot(pred_data, aes(x = user_type, y = predicted, fill = day_label)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("#619CFF", "#F8766D"), name = "Day Type") +
  labs(
    title    = "Predicted Mean Hourly Rentals",
    subtitle = "Bar chart",
    x        = "User Type",
    y        = "Predicted Mean Rentals"
  ) +
  theme_calc()

# --- Plot 2: Interaction plot ---
p2 <- ggplot(pred_data, aes(x = day_label, y = predicted,
                             color = user_type, group = user_type)) +
  geom_point(size = 4) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c("casual" = "#F8766D", "registered" = "#619CFF"),
                     name = "User Type") +
  labs(
    title    = "Interaction: Workday × User Type",
    subtitle = "Crossing lines show reversed effect",
    x        = "Day Type",
    y        = "Predicted Mean Rentals"
  ) +
  theme_calc()

# --- Combined ---
p1 + p2 +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

This suggests that more registered users might use bike sharing more on workdays (for commute/exercise) while casual users might use the service for recreation on holidays/weekends.

The Analysis of Deviance Table shows including the interaction term, including it reduces the residual deviance from the reduced model by 151,556, providing strong evidence that workday status affects casual and registered users differently. We can thus say that there is a practically meaningful difference in relationships registered/casual bike users have with working days.

### (RQ.4)

Our estimated `day_of_season` coefficients were statistically significant, since `day_of_season` and `day_of_season:user_typeregistered` had individual p-values of
```{r}
usertype_coefs = c("day_of_season", "day_of_season:user_typeregistered")
s$coefficients[usertype_coefs, "Pr(>|z|)"]
```

and a p-value when added together
```{r}
no_day_of_season_model = glm(rentals ~ (temp + windspeed + hum + weathersit + workingday + season)*user_type, bike_transformed, family = poisson())

anova(no_day_of_season_model, model)[2, "Pr(>Chi)"]
```

with individual values
```{r}
coef(model)[usertype_coefs]
```

These indicate that there was a positive relationship between the day of the season and average bike rentals. Under this, bike rentals increased in later parts of seasons for all users - though to a slightly greater extent for registered users. Specifically, it is estimated that for every day that passed in a season, the estimated mean number of casual bike rentals in an hour increased by a factor of 
```{r}
coef(model)[["day_of_season"]] |> exp()
```
and for registered users
```{r}
coef(model)[usertype_coefs] |> sum() |> exp()
```

However, these values are very small, especially since this effect prorated over an entire season (about 90 days) would only result in a multiplicative increase in bike rentals by 
```{r}
(coef(model)[usertype_coefs] |> sum() |> exp())^90
```

which is a fairly small increase. Therefore, even though the coefficient is statistically significant, it would seem that there isn't a practically significant effect on rentals, with weather and `user_type` having a much greater effect on the number of rentals.

## Conclusion

<!-- Briefly summarize your main findings and their implications. -->

Bike rentals are significantly associated with different climate/weather conditions. They

- Significantly increased for higher temperatures
- Increased for non-working days

In addition, registered users tended to behave differently from casual users, with registered users often having a dampened or even reversed relationship between weather predictors and working days. Registered users were generally less affected by the weather, and saw an increase in rentals on working days compared to the decrease seen in casual users.

We hypothesized that bike rentals increased as a season went along (later parts of a season were associated with more rentals), but we decide that there isn't a significant relationship.

## Limitations

- Poisson regression assumes mean equals variance, but our residual deviance of 2,177,102 on 34,734 degrees of freedom suggests substantial over dispersion. A negative binomial/quasi-Poisson model may thus be worth exploring in this case.

- The diagnostic plots reveal the model systematically overestimates rentals at high counts, suggesting the linear predictor may not fully capture extreme values. Which might suggest that a transform on one of predictors might be suitable. 

- Hours within the same day are likely correlated (Temporal dependence), violating the independence assumption of Poisson regression. Addressing this would require models that address the time series nature of the data.


## References

<!-- Cite the data source and any articles or resources you used. -->

::: {#refs}
:::

## Appendix {.appendix}

<!-- Optional: include additional code, tables, or figures here.
This section does not count toward the page limit. -->

The model was fit using

```{r, echo=TRUE, eval=FALSE, ref.label="model_fit"}
#| code-overflow: wrap
```


Analysis of Deviance Table (to check if the Interaction effect between Working day and User Type)

```{r}
#| code-overflow: wrap
model_no_wday_interaction <- glm(
  rentals ~ (temp + windspeed + hum + weathersit + day_of_season + season) * user_type 
          + workingday + user_type,
  bike_transformed, 
  family = poisson()
)

anova(model_no_wday_interaction, model, test = "LRT")
```


Expanded Predictor Information

Dataset predictors we used are described below

| Predictor | Description |
| -- | ------ |
| `temp` | The normalized temperature of the hour (unitless), in $\left[0, 1\right]$. Found by taking the temperature in $C^{\circ}$ and dividing by 41 |
| `windspeed` | The normalized windspeed of a day (unitless), in $\left[0, 1\right]$. Found by taking the windspeed of the hour (unit is unstated in dataset) and dividing by 41 |
| `hum` | The normalized humidity (unitless), in $\left[0, 1\right]$; taken from original humidity and then dividing by 100 |
| `weathersit` | Factor representing the amount of precipitation/cloud cover. In $\{1, 2, 3, 4\}$, where <br> **1:** Clear, Few clouds, Partly cloudy, Partly cloudy <br> **2:** Mist + Cloudy, Mist + Broken clouds, Mist + Few clouds, Mist <br> **3:** Light Snow, Light Rain + Thunderstorm + Scattered clouds, Light Rain + Scattered clouds <br> **4:** Heavy Rain + Ice Pallets + Thunderstorm + Mist, Snow + Fog |
| `season` | Indicator for what season it is. In the data, is an integer from 1-4, where 1 is Spring, 2 Summer, 3 Fall, and 4 Winter. For our model, we transform it into categorical data where `Spr` is for Spring, `Sum` is for Summer, `Aut` is for Fall, and `Win` is for Winter |
| `workingday`| Indicator of if day was a workday or not (1 if workday, 0 if not) |

We also create new predictors from the dataset, described below

| Predictor | Description |
| -- | ------ |
| `user_type` | Indicator of if the number of rentals is taken from the `casual` or `registered` columns of the dataset (0 if `casual`, 1 if `registered`). We reshaped the data to be longer: instead of each row containing both a casual and registered count, we separated them out into two rows, with the casual row having `user_type` = 0, and 1 for the registered row |
| Day of season | The day of the season (e.g. 0 for the first day of Spring, 1 for the second, etc..., then 0 for the first day of Fall, 1 for the second, etc...). |
